
\begin{thebibliography}{}
    \bibitem{GelMan}S.I.~Gelfand and Y.I.~Manin, ``Homological algebra'', 2nd Ed., Springer, 2003.
    \bibitem{CTCS} B.~Milewski, ``\href{https://github.com/hmemcpy/milewski-ctfp-pdf/}{Category Theory for Programmers}''.
    \bibitem{JetNest} Jet Nestruev, ``Smooth Manifolds and Observables'', Springer, 2003.
    \bibitem{Connes} Connes and Marcolli, ``Noncommutative Geometry, Quantum Fields and Motives''.
    \bibitem{compact.proof} \url{https://proofwiki.org/wiki/Topological_Product_of_Compact_Spaces}.
    \bibitem{Schuller} \href{https://www.youtube.com/playlist?list=PLPH7f_7ZlzxTi6kS4vCmv4ZKm9u8g5yic}{F. Schuller's video lectures on Differential Geometry and Physics}.
    
    \bibitem{Lee}J.M.~Lee, ``Introduction to Smooth Manifolds'', Springer, 2nd Ed., 2013.
    
    \bibitem{LeeTop}J.M.~Lee, ``Introduction to Topological Manifolds'', Springer, 2nd Ed., 2011. 
    
    \bibitem{Gadea} P.M.~Gadea et al., ``Analysis and Algebra on Differentiable Manifolds: A Workbook for Students and Teachers'', Springer, 2nd Ed., 2013.
    
    \bibitem{LieRiem}J.M.~Lie, ``Riemannian manifolds'', Springer, 1st Ed., 1997.
    
    \bibitem{Jost} J.~Jost, ``Riemannian Geometry and Geometric Analysis'', Springer, 6th Ed., 2011.
    
    
    \bibitem{smooth.structures} \url{https://mathoverflow.net/questions/47569/what-makes-four-dimensions-special}.
    
    \bibitem{Aguilar} Marcelo A.~Aguilar and Carlos Prieto, \href{https://paginas.matem.unam.mx/cprieto/phocadownloadpap/fiber%20bundles.pdf}{``Fiber Bundles''}, 2010.
    \bibitem{Dundas} Bj{\o}rn Ian Dundas, \href{https://citeseerx.ist.psu.edu/document?repid=rep1&type=pdf&doi=0603e3854c4dc13a2a283072ef834b05e8eec607}{``Differential Topology''}, 2013.
    \bibitem{hopf} Hopf fibration visualization, \url{https://www.youtube.com/watch?v=AKotMPGFJYk}.
    \bibitem{RS1} G.~Rudolph and M.~Schmidt, ``Differential Geometry and Mathematical Physics, Part I'', Springer, 2013.
    \bibitem{RS2} G.~Rudolph and M.~Schmidt, ``Differential Geometry and Mathematical Physics, Part II'', Springer, 2017.
    \bibitem{Atiyah} M.F.~Atiyah, ``Vector Bundles over an Elliptic Curve'', Proc.London Math. Soc. \textbf{3} 7, 1957, 414-452.
    
    \bibitem{Bredon} Glen E.~Bredon, ``Topology and Geometry'', 1993.
    
    
    \bibitem{Milnor}J.~Milnor, ``Morse Theory'', Princeton University Press, 1973.
    
    \bibitem{Munk}J. R.~Munkres, ``Topology''.
    
    \bibitem{Hatcher} A.~Hatcher, ``Algebraic Topology'', 2002.
    \bibitem{HatcherVB} A.~Hatcher, \href{https://pi.math.cornell.edu/~hatcher/VBKT/VB.pdf}{``Vector Bundles and K-Theory''}, draft.
    \bibitem{Huybrechts} D.~Huybrechts, ``Complex Geometry: An Introduction'', Springer, 2005.
    \bibitem{BottTu} R.~Bott and L.W.~Tu, ``Differential Forms in Algebraic Topology'', Springer, 1982.
    \bibitem{tomDieck} T.~tom Dieck, ``Algebraic Topology'', AMS, 2008.
    \bibitem{Rotman} J.J.~Rotman, ``An introduction to Homological Algebra'', 2nd Ed., Springer, 2009.
    \bibitem{Bergman} G.M.~Bergman, \href{https://math.berkeley.edu/~gbergman/245/}{``An Invitation to General Algebra and Universal Constructions''}, 2020 (also published by Springer).
    \bibitem{Leinster} T.~Leinster, \href{https://arxiv.org/abs/1612.09375}{``Basic Category Theory''}, 2016.
    \bibitem{Perrone} P.~Perrone, \href{https://arxiv.org/abs/1912.10642}{``Notes on Category Theory with examples from basic mathematics''}, 2019.
    \bibitem{May} J.P.~May, \href{https://www.math.uchicago.edu/~may/CONCISE/ConciseRevised.pdf}{``A Concise Course in Algebraic Topology''}, 1999.
    \bibitem{Spanier} E.H.~Spanier, ``Algebraic Topology''. Springer, New York, 1994.
    \bibitem{DK} J.J.~Duistermaat and J.A.C.~Kolk, ``Lie Groups'', Springer, Berlin, 2000.
    \bibitem{Neeb} K.-H.~Neeb, \href{https://cel.hal.science/cel-00391789}{``Infinite-Dimensional Lie Groups''}, HAL, 2009.
    \bibitem{Warner} F.-W.~Warner, ``Foundations of Differentiable Manifolds and Lie Groups'', Springer, 1971.
    \bibitem{Onishchik} A.L.~Onishchik and E.B.~Vinberg, ``Lie Groups and Lie Algebras I'', Springer, 1993.
    \bibitem{Sepanski} M.R.~Sepanski, ``Compact Lie Groups'', Springer, 2007.
    \bibitem{Munkres} J.R.~Munkres, ``Elements of algebraic topology'', 1984, Taylor\&Francis Group.
    \bibitem{Shastri} A.R.~Shastri, ``Basic algebraic topology'', 2014, CRC Press.
    \bibitem{HN} J.~Hilgert, K.H.~Neeb, ``Structure and Geometry of Lie Groups'', Springer, 2010.
    \bibitem{Sharpe}  R.W.~Sharpe, S.S.~Chern, ``Differential Geometry: Cartan's Generalization of Klein's Erlangen Program'', 1997.
    \bibitem{Cap} A.~Cap, J.~Slov\'ak, ``Parabolic Geometries I'', 2009.
    \bibitem{Kolar} I.~Kol\'a\v{r}, P.W.~Michor, J.~Slov\'ak, ``Natural operations in differential geometry'', 1993.
    \bibitem{Vakar} M.~V\'ak\'ar, ``Principal Bundles and Gauge Theories'', 2011.
    \bibitem{Hoyo} M.~del~Hoyo, \href{https://arxiv.org/abs/1512.03847}{``Complete connections on fiber bundles''}, Indagationes Mathematicae (\textbf{27} 4), 2015.
    \bibitem{Kobayashi} S.~Kobayashi, ``Differential Geometry of Complex Vector Bundles'', Princeton University Press, Princeton, 1987.
    \bibitem{Spivak2} M.~Spivak, ``A comprehensive introduction to differential geometry'', vol.~2, 1999.
    \bibitem{OFarrill} J.~Figueroa-O'Farrill, \href{https://arxiv.org/abs/2009.01948v1}{``On the intrinsic torsion of spacetime structures''}, 2020.
    
    
    
\end{thebibliography}